\documentclass[10pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}

\title{How Much Adaptation Is Enough?\\Drift Geometry and Minimal Calibration in Brain-to-Text BCI}
\author{AdaptLadder-BCI}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Brain-to-text BCI decoders are vulnerable to cross-session changes in neural feature
distributions. This short paper probes how much input adaptation is needed when a
pretrained brain-to-text decoder is forced across sessions. Using the official T15
GRU baseline, we show that native day-specific input layers reproduce low validation
phoneme error, while non-native input layers cause large decoder-facing degradation.
Simple moment corrections fail to recover this loss, learned small-calibration
adapters recover only a modest fraction, and recovery depends on drift geometry.
\end{abstract}

\section{Introduction}
Cross-session drift can degrade BCI decoder reliability and increase recalibration
burden. Rather than immediately changing decoder weights, we ask how far simple
input-space corrections can go, and when the geometry of the drift suggests that
stronger adaptation is needed.

\section{Related Work}
To be filled with brain-to-text BCI, neural decoder adaptation, test-time adaptation,
and recalibration references.

\section{Data}
Primary analysis uses T15 copy-task neural features. The first sanity pass reports
session counts, trial counts, trial lengths, feature dimensionality, and available
labels/splits.

\section{Methods}
\subsection{Decoder-facing stress test}
We evaluate the official T15 GRU checkpoint on the validation split using greedy CTC
phoneme decoding. The native-day condition sends each session through its own learned
day-specific input layer. The cross-day condition sends all target sessions through a
fixed non-native source input layer before the frozen GRU and output head.

\subsection{Adaptation ladder}
We compare no correction, target-session z-scoring, source moment matching,
supervised diagonal affine calibration, and supervised input-layer calibration.
Calibration methods use the first $K$ labeled validation trials from each target
session and are evaluated only on the remaining trials.

\subsection{Drift geometry}
For each target session relative to the middle source session, we estimate temporal
distance, mean and scale shifts, covariance/CORAL-style distances, and PCA subspace
distances. We correlate these metrics with cross-day PER and with the recovered
fraction of the native-day performance gap.

\subsection{Geometry-guided source selection}
We also test whether drift geometry can be used constructively. For each target
session, we compute geometry distances to candidate non-native source sessions and
select the nearest source input layer before running the frozen GRU. We evaluate an
offline setting with all other sessions as candidates and a stricter past-only
setting that excludes future source sessions.

\subsection{Beginning-of-day K-shot selection}
To approximate deployment, we estimate target-session geometry from only the first
$K$ validation trials, without labels. We then select the nearest past source input
layer and evaluate decoding only on the remaining trials from that target session.
We also compare this rule against a previous-session baseline that uses the
immediately preceding input layer.

\subsection{Optional T12 geometry feasibility}
We prepare an optional T12 validation using the diagnostic blocks from the Dryad
release for the high-performance speech neuroprosthesis dataset. This analysis is
geometry-only: it does not attempt to reproduce the T12 decoder. The goal is to
test whether diagnostic block neural features show a similar across-day geometry
structure, including whether previous-session recency is a strong source
candidate and whether geometry occasionally selects older sessions.

\section{Results}
\subsection{Native-day decoding is reproduced}
The official checkpoint reaches 9.06\% phoneme-weighted PER on the T15 validation
split in the native-day condition, confirming that the local probe reproduces the
expected decoder behavior.

\subsection{Cross-day input mismatch breaks the decoder}
Forcing all sessions through non-native input layers raises phoneme-weighted PER to
22.67--34.43\% across middle, early, and late source choices. The middle source
`t15.2023.11.26' gives 22.67\% PER, while early and late sources give 28.81\% and
34.43\%. All three source choices harm 40/41 validation sessions.

\subsection{Moment correction is not enough}
Using the middle source, no correction gives 22.67\% PER. Target z-scoring gives
22.75\% PER and source moment matching gives 22.73\% PER. Thus full-session
unsupervised mean/variance alignment does not recover the cross-day loss.

\subsection{Small learned calibration recovers only a modest fraction}
With first-$K$ calibration and five training epochs, diagonal affine calibration
improves weighted PER from 21.51\% to 20.21\% at $K=20$, recovering about 10.1\% of
the native-day gap. Full input-layer calibration gives a similar 20.27\% PER at
$K=20$, recovering about 9.7\%. These gains are consistent but modest.

\subsection{Drift geometry predicts degradation and recovery}
Temporal and covariance/CORAL-style distances from the source strongly predict
cross-day PER. For example, cross-day PER has Spearman $\rho \approx 0.86$--0.87
with absolute days and relative covariance shift at $K=5/10$ subsets
(permutation $p=0.001$), and $\rho \approx 0.77$ at $K=20$. Recovery is weaker and
more variable, but input-layer recovery at $K=5$ decreases with distance
($\rho=-0.49$ for absolute days, $p=0.003$; $\rho=-0.43$ for covariance shift,
$p=0.009$). A covariance-distance median split shows higher input-layer recovery
for near sessions at $K=5$ and $K=20$, while $K=10$ is mixed.

\subsection{Geometry can select a better non-native input layer}
Selecting the nearest non-native source input layer by relative covariance distance
substantially reduces cross-day degradation without training new parameters. In the
offline all-candidate setting, geometry-nearest selection reaches 11.38\%
phoneme-weighted PER, compared with 22.67\% for the fixed middle source and 9.06\%
for native-day decoding. It improves 39/41 sessions relative to the fixed middle
source. In the stricter past-only setting, geometry-nearest selection reaches
13.58\% PER on 40 evaluable sessions, compared with 22.76\% for the fixed middle
source on the same sessions and 9.10\% native-day PER. This suggests that drift
geometry can guide source-layer selection, not only post-hoc analysis of failures.

\subsection{A short unlabeled calibration window is enough for source selection}
Beginning-of-day K-shot selection preserves most of the geometry-selection benefit
while using only a short unlabeled target-session window. With past-only source
candidates, K-shot geometry selection reaches 14.00\% PER at $K=5$, 13.19\% at
$K=10$, and 12.74\% at $K=20$, evaluated only on the remaining trials. On the same
subsets, the fixed middle source gives 22.51\%, 22.40\%, and 22.30\%, while
native-day decoding gives 8.95\%, 8.88\%, and 8.90\%. This corresponds to recovery
of 62.8\%, 68.1\%, and 71.3\% of the fixed-source gap, improving 36/41, 36/40,
and 34/37 sessions relative to fixed middle.

\subsection{Previous-session recency is a strong baseline}
We further ask whether K-shot geometry selection is more than a recency heuristic.
Using the immediately previous input layer gives 13.09\% PER at $K=5$, 12.84\%
at $K=10$, and 12.21\% at $K=20$ on the same remaining-trial subsets, slightly
better than K-shot geometry selection. Geometry selects the previous session in
30/41, 31/40, and 26/37 target sessions for $K=5/10/20$. When it selects an
older source, it improves over the previous-session source in 3, 3, and 4
sessions, respectively, but the aggregate result still favors the previous
session. Thus temporal recency is a strong baseline; geometry currently appears
most useful as a way to detect possible non-recent source matches rather than as
a replacement for the previous-session rule.

\subsection{A simple geometry override is not enough}
We next test a recency-aware gate: use the previous-session input layer by
default, and switch to the geometry-selected older source only when its
covariance distance is less than $\alpha$ times the distance to the previous
source. Sweeping $\alpha \in \{0.5,0.6,0.7,0.8,0.9,1.0\}$ shows that this single
ratio gate does not yet produce a meaningful improvement over recency. At
$\alpha=0.9$, the override ties previous-session decoding at $K=5$, is slightly
worse at $K=10$ (12.90\% vs 12.84\%), and gives only a very small gain at
$K=20$ (12.18\% vs 12.21\%). At $\alpha=1.0$, the rule approaches unconditional
geometry selection and becomes worse. Thus geometry remains useful as a candidate
override signal, but the gate needs richer confidence features than a single
distance ratio.

\section{Discussion}
The central question is whether lightweight correction is enough, and when the
remaining error calls for stronger adaptation. These preliminary results suggest that
moment-based correction is too weak, small learned adapters recover some loss, and
drift geometry may help choose the next rung of the adaptation ladder. The strongest
current result is geometry-guided source selection: before learning any new adapter,
choosing a geometrically compatible existing input layer recovers much of the loss
from a fixed non-native source. Importantly, this does not require full-session
access: an unlabeled beginning-of-day window of 20 trials recovers about 71\% of the
fixed-source loss on the rest of the session. However, the previous-session input
layer is an even stronger simple baseline in this K-shot analysis. Future
geometry-aware gates should therefore be compared against recency and may be most
valuable when they identify sessions whose neural geometry returns to an older
input-layer state. A first distance-ratio override gate yields only a tiny gain at
$K=20$, so the immediate research target is not another adapter but a better
confidence model for deciding when to override recency.

\section{Limitations}
Distribution alignment alone is not a sufficient performance metric. WER, PER, CTC
loss, decoder confidence, or related decoder-facing metrics are required for a stronger
paper. T12 validation is currently scaffolded but should not be interpreted as a
replication result until the diagnostic blocks are downloaded locally and the
geometry-only analysis is run on real T12 MAT files.

\section{Conclusion}
Adaptation ladders can provide a practical way to decide how much correction a
brain-to-text BCI session needs before moving to heavier recalibration. The current
T15 results support a geometry-aware version of this ladder: easy sessions may need
only small input calibration, while geometrically distant sessions likely require a
stronger adapter or renewed supervised recalibration.

\end{document}
